Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-04-02 agujournaltemplate.tex
Update Significance Levels, add comments

% Figure 2
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. The patterns emerging from the -u 10m wind component, the temperature and the directly related \qty{21}{\celsius} isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m. These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes. The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration. Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.

% Figure 3
To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between seasons. The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters. In some years, particularly during 2000, 2005, 2007, 2008 and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter. Genus richness is consistently highest in the top \qty{25}{meters} depth interval, which also shows the most variability over the time series. Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011. The collapse in diversity following the start of the time series does not follow the signal shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling this genus diversity by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016. The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$), yet there is a consistent pattern of higher Shannon diversity at lower depths for all seasons except for Fall. 
Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$). The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

% Figure 4
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. 
Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, the turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. 
To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. Upwelling related variables show distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Artboard 1.pdf}
\caption{Map of CARIACO Time Series location in the Cariaco basin in Venezuela in the Caribbean Ocean, located at \ang{10.50}N, \ang{64.67}W}
\label{fig:map}
\end{figure}

@@

\begin{figure}
\begin{center}
\noindent\includegraphics[width=480pt]{fig/DepthDivTimeSeries_updated.pdf}
\end{center}
\caption{Time series and seasonality of chlorophyll and diversity indices across four depth intervals. (a,c,e,g) Annual mean across time series for years with more than 5 monthly sampling. Transparent dots show monthly raw values. (b,d,f,h) Boxplots of monthly measurements across four depth intervals grouped by season.}
\label{fig:divts}
\end{figure}

@@

\begin{table}
\caption{Wilcoxon sum rank test with continuity correction for variables between the two clusters. Values for in-situ data are interpolated across the top 100 meters for each individual monthly cruise sampling.}
\label{table:clustering}
\centering
\begin{tabular}[t]{lrrl}
\toprule
Variable & W & p.value & Significance\\
\midrule
AMO & 3494.0 & 1.86e-14 & ***\\
MEI v.2 & 9523.5 & 0.006 & ***\\
u-component 10 m wind speed & 7074.0 & 0.143 & n.s.\\
Isotherm \qty{21}{\celsius} & 5293.0 & 0.034 & **\\
\addlinespace
Temperature & 4763.0 & 0.003 & ***\\
$NO_3$ & 6378.0 & 0.018 & **\\
$PO_4$ & 7359.0 & 0.012 & **\\
$SiO_4$ & 5998.0 & 0.736 & n.s.\\
Salinity & 7636.0 & 0.002 & ***\\
\addlinespace
Genus Richness & 5913.0 & 0.954 & n.s.\\
Shannon Index & 5142.0 & 0.108 & n.s.\\
Pielou Index & 4874.0 & 0.029 & **\\
\bottomrule
\end{tabular}
\end{table}

@@

>>>>

% Figure 2
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. We included marks to highlight those years with missing monthly measurements, as these make the annual means less reliable, but they generally do not affect the general patterns observed, even if substituted by monthly means across the time series. Particularly towards the alter years of the time series, the sampling became sparse. The patterns emerging from the -u 10m wind component, the temperature and the directly related \qty{21}{\celsius} isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m. These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes. The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration. Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.

% Figure 3
To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between the rainy and upwelling seasons. The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters. In some years, particularly during 2000, 2005, 2007, 2008 and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter. Genus richness is consistently highest in the top \qty{25}{meters} depth interval, which also shows the most variability over the time series. Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011. The collapse in diversity following the start of the time series does not follow the signal shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling this genus diversity by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016. The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$), yet there is a consistent pattern of higher Shannon diversity at lower depths for all seasons except for Fall. 
Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$). The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

% Figure 4
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. 
Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, the turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. 
To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. Upwelling related variables show distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
\caption{Map of CARIACO Time Series location in the Cariaco basin in Venezuela in the Caribbean Ocean, located at \ang{10.50}N, \ang{64.67}W}
\label{fig:map}
\end{figure}

@@

\begin{figure}
\begin{center}
\noindent\includegraphics[width=480pt]{fig/DepthDivTimeSeries_updated.pdf}
\end{center}
\caption{Time series and seasonality of chlorophyll \textit{a} and diversity indices across four depth intervals. (a,c,e,g) Annual mean across time series for years with more than 5 monthly sampling. Transparent dots show monthly raw values. (b,d,f,h) Boxplots of monthly measurements across four depth intervals grouped by season. Upwelling season lasts from December to April, Rainy Season from May to November.}
\label{fig:divts}
\end{figure}

@@

\begin{table}
\caption{Wilcoxon sum rank test with continuity correction for variables between the two clusters. Values for in-situ data are interpolated across the top 100 meters for each individual monthly cruise sampling.}
\label{table:clustering}
\centering
\begin{tabular}[t]{lrrl}
\toprule
Variable & W & p.value & Significance\\
\midrule
AMO & 3494.0 & 1.86e-14 & ***\\
MEI v.2 & 9523.5 & 0.006 & **\\
u-component 10 m wind speed & 7074.0 & 0.143 & n.s.\\
Isotherm \qty{21}{\celsius} & 5293.0 & 0.034 & *\\
\addlinespace
Temperature & 4763.0 & 0.003 & **\\
$NO_3$ & 6378.0 & 0.018 & *\\
$PO_4$ & 7359.0 & 0.012 & *\\
$SiO_4$ & 5998.0 & 0.736 & n.s.\\
Salinity & 7636.0 & 0.002 & **\\
\addlinespace
Genus Richness & 5913.0 & 0.954 & n.s.\\
Shannon Index & 5142.0 & 0.108 & n.s.\\
Pielou Index & 4874.0 & 0.029 & *\\
\bottomrule
\end{tabular}
\end{table}

@@

% FROM FRAKER ET AL. 2022
%"Note that although this analysis identifies predictive relationships between drivers and the biological indicators, these relationships do not necessarily imply mechanistic relationships. As such, this analysis represents the first stages of an overall analysis (a narrowing of the potential key relationships and when they matter from a much larger dataset), and additional statistical analyses, modeling, and experimental work may be required to develop a full mechanistic understanding of driver-response relationships and ecosystem change."
